\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-23 paper/algorithm.tex
fix: info loss => info lost, and adjusted small discussion

In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the amount of information loss that a user is willing to tolerate relative to the minimum information loss possible given the imputation operations.\todo{Get rid of discussion of information loss. It's a technical term and we don't really tackle it.}

@@

Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

>>>>

In order to pick a final plan from the frontier, our model introduces a parameter $\alpha$ which encodes an upper bound on the $\textsc{Loss}$ value that the user is willing
to tolerate relative to the minimum possible amongst frontier plans.

@@

Using the drop operator incurs a loss of information of 1 (i.e. we lose all information that
the dropped tuples might have encoded). Imputing null fields incurs a loss penalty $p \in (0, 1]$,
which is inversely proportional to the number of total cells available to train the imputation algorithm. 
Total cells are defined as the product of the number of attributes and the number of tuples.
Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information lost \textit{decreases} with more values to work
with. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

### 2017-02-23 paper/algorithms/plan-toplevel.tex
fix: info loss => info lost, and adjusted small discussion

% \begin{algorithm}
\begin{algorithmic}[1]
  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information loss}
  \Ensure{An optimized query plan.}

>>>>

% \begin{algorithm}
\begin{algorithmic}[1]
  \Require{A set of tables $T$, a relation $F: T \times \Phi$ between tables and filter predicates, a relation $J : T \times \Psi \times T$ between tables and join predicates, a set of projection attributes $P$,  an optional set of grouping attributes $G$ and aggregator function $A$, and a parameter $\alpha \in [0, 1]$ that expresses trade-off performance and information lost}
  \Ensure{An optimized query plan.}

### 2017-02-23 paper/conclusion.tex
fix: info loss => info lost, and adjusted small discussion

\section{Conclusion}
As shown in \ProjectName{}, missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
information loss and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC} and \textit{freeCodeCamp} survey-based datasets
showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute in a fraction of the time compared to the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

>>>>

\section{Conclusion}
As shown in \ProjectName{}, missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
information lost and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC} and \textit{freeCodeCamp} survey-based datasets
showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute in a fraction of the time compared to the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be small in all queries considered. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

### 2017-02-23 paper/relatedwork.tex
fix: info loss => info lost, and adjusted small discussion

In \cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information loss and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

In contrast to this work, \ProjectName{} is targeted at generic value imputation, not necessarily tailored to 
time-series. The optimizer is not based on equivalence transformations, nor are there guarantees of equal
results under different conditions. Instead, the optimizer allows users to pick their tradeoff between
runtime cost and information loss. The search space considered by our optimizer is broader, not just
forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan
(with some restrictions). The novelty of our contribution lies in the successful incorporation of
imputation operations in non-trivial query plans with cost-based optimization.

>>>>

In \cite{akande2015empirical}, Akande et al analyze the performance of various
multiple imputation techniques on the American Community Survey dataset. 
The computational difficulties of imputing on large base
tables are well known and can limit approaches. For example, Akande finds that
one approach (MI-GLM) is prohibitively expensive when attempting to impute on data that
includes variables with potentially large domains (ten categories in their case).
In contrast, \ProjectName{} allows users to specify a tradeoff between
information lost and runtime performance, allowing users to perform queries
directly on the entire dataset. Furthermore,
the query planner's imputation is guided by the requirements of each specific
query's operators, rather than requiring broad assumptions about query
workloads.

@@

In contrast to this work, \ProjectName{} is targeted at generic value imputation, not necessarily tailored to 
time-series. The optimizer is not based on equivalence transformations, nor are there guarantees of equal
results under different conditions. Instead, the optimizer allows users to pick their tradeoff between
runtime cost and information lost. The search space considered by our optimizer is broader, not just
forecast-first/forecast-last plans, but rather imputation operators can be placed anywhere in the query plan
(with some restrictions). The novelty of our contribution lies in the successful incorporation of
imputation operations in non-trivial query plans with cost-based optimization.
